Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package \vskip15pt Supplementary Materials

### 2018-09-28 docs/MEE_submission/manuscript_MEE.tex
updated MS

In the fields of ecology and evolution, researchers make use of data presented in primary literature for comparative and meta-analyses. These techniques rely on descriptive statistics (e.g. means, standard deviations (SD), sample sizes, correlation coefficients) extracted from primary literature. As well as being presented in the text or tables of research papers, descriptive statistics are frequently presented in figures and so need to be manually extracted using digitising programs.

Although there are several tools that extract data from figures (e.g. \proglang{DataThief} \citep{DataThief}, \proglang{GraphClick} \citep{GraphClick}, \proglang{WebPlotDigitizer} \citep{WebPlotDigitizer}, see Table \ref{tab:comparison}), these tools do not cater to needs of meta-analysts for four main reasons (here we focus on meta-analysis, although many  points apply to extraction for comparative analysis). First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}. Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later using different software. Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images and export the resulting digitised data into individual files one-by-one. These data often subsequently need to be imported and edited using different software. Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently, a large amount of additional data manipulation is required, that is different across plots types. For example, data are frequently presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented.

Data extraction from figures is therefore an incredibly time-consuming process as existing software does not provide an optimized, reproducible research pipeline to facilitate data extraction and editing. Here, we present an interactive \proglang{R} package, \pkg{metaDigitise} (available at https://github.com/daniel1noble/metaDigitise), which is designed for large scale, reproducible data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools to extract data from common plot types (mean/error plots, box plots, scatter plots and histograms, see Figure \ref{fig:all_extract}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. The necessary calculations are carried out on calibrated data immediately after extraction so that comparable descriptive statistics can be obtained quickly. Summary data from multiple figures is returned into a single data frame which can be can easily exported or used in downstream analysis within \proglang{R}. Completed digitisations are automatically saved for each figure, meaning users can redraw their digitisations (along with metadata) on figures, make corrections and access calibration and processed (i.e., summarised) data. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analyses to be updated more efficiently.

@@

\section{\pkg{metaDigitise} and Reproducibility}
The \pkg{metaDigitise} package has one main function, \fct{metaDigitise}, which interactively takes the user through the process of extracting data from figures (see Supplementary Material S1 for a full tutorial). \fct{metaDigitise} works on a directory containing images of figures copied from primary literature, in .png, .jpg, .tiff, .pdf format, specified to \fct{metaDigitise} through the \code{dir} argument. \fct{metaDigitise} recognizes all the images in the given directory and automatically imports them one-by-one, allowing the user to extract the relevant information about a figure as they go. Figures can be organised in different ways for a project, but we would recommend having all figures for one project in a single directory with an informative and unambiguous naming scheme (e.g. paper\_figure\_trait.png). This expedites digitisation by preventing users from having to constantly change directories and / or open new images.

@@

The complete digitisation process can then be reproduced at a later stage, shared with collaborators and presented as supplementary materials for a publication, regardless of the computer it is run on. For an analysis to be updated, new figures can simply be added to the directory and \fct{metaDigitise} run to incorporate the new data.

@@

\section{Image Processing}
Running \fct{metaDigitise} presents the user with three options; `Process new images', `Import existing data' or `Edit existing data', which can be used during and after digitisation to execute a range of functions (see Figure \ref{fig:all_extract} – 'Editing' and 'Importing' are discussed in the next section). Selecting `Process New Images' starts the digitisation process on images within the directory that have not previously been digitised. For all plot types, \fct{metaDigitise} requires the user to calibrate the axes in the figure, by clicking on two known points on the axis in question, and entering the value of those points (Figure \ref{fig:all_extract}). \fct{metaDigitise} then calculates the value of any clicked points in terms of the figure axes. This is based on the calibration used in the \pkg{digitize} R package \citep{Poisot2011}. For mean/error and box plots, only the y-axis is calibrated (Figure \ref{fig:all_extract}), assuming the x-axis is redundant. For scatter plots and histograms both axes are calibrated (Figure \ref{fig:all_extract}).

As figures may have been copied from older, scanned publications, they may not be perfectly orientated. This makes calibration of the points in the figure problematic. \fct{metaDigitise} allows users to rotate the image (Figure S2A,B). Furthermore, mean/error plots, box plots and histograms, may be presented with horizontal bars. \fct{metaDigitise} assumes that bars are vertical, but allows the user to flip the image to make the bars are vertical (Figure S2C,D). \pkg{metaDigitise} also allows back calculation of data presented on log axes.

@@

\section{Importing and Editing Previously Digitised data}
\pkg{metaDigitise} is also able to re-import, edit and re-plot previously digitised figures. When running \fct{metaDigitise}, the user can choose to 'Import existing data', which returns previously digitised data, from a single figure or all figures. Alternately, the \fct{getExtracted} function returns the data from previous digitisations, but without user interaction, allowing easier integration into larger scripts. 'Edit existing data' allows the user to re-plot or edit information for digitisations that have previously be done. Re-plotting digitisations with all metadata is an important reproducibility feature, as it allows users to see exactly what information has been extracted, as well as making it easy to spot and data extraction errors.

@@

\section{Software Validation}
In order to evaluate the consistency of digitisation with \pkg{metaDigitise} between users, fourteen people digitized sets of 14 identical images created from a simulated dataset (see supplements). We found no evidence for any inter-observer variability in digitisations for the mean (ICC = 0, 95\% CI = 0 to 0.029, \textit{p} = 1), SD (ICC = 0, 95\% CI = 0 to 0.033, \textit{p} = 0.5) or correlation coefficient (ICC = 0.053, 95\% CI = 0 to 0.296, \textit{p} = 0.377). There was little bias between digitised and true values, on average 1.63\% (mean = 0.02\%, SD = 4.9\%, \textit{r} = -0.03\%) and there were small absolute differences between digitised and true values, on average 2.18\% (mean = 0.40\%, SD = 5.81\%, \textit{r} = 0.33\%) across all three descriptive statistics. SD estimates from digitisations are clearly most error prone. The mean absolute differences for each plot type clearly show that this effect is driven by extraction from box plots and histograms (\% difference; box plot: 15.805, histogram: 5.210, mean/error: 1.500, scatter plot: 0.433). SD estimation from box plot descriptive statistics is known to be more error prone, especially at small sample sizes \citep{Wan2014}.

We also used simulated data to test the accuracy of digitisations with respect to known values (see supplements). \pkg{metaDigitise} was extremely accurate at matching clicked points to their true values essentially being perfectly correlated with the true simulated data for both the \textit{x}-variable (Pearson's correlation; $r > 0.999$, \textit{t} = 2137.4, \textit{df} = 78, $p < 0.001$) and \textit{y}-variable ($r > 0.999$, \textit{t} = 1897.8, \textit{df} = 78, $p < 0.001$) in scatterplots.

@@

In contrast to some other packages \pkg{metaDigitise} does not extract lines from figures. Line extraction is not particularly useful for most comparative or meta-analytic work, although we recognise that it may be useful in fields other than these. Should a user like to extract lines with \pkg{metaDigitise}, we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points (accessed by choosing to return calibrated rather than summary data) to estimate the parameters needed.

@@

\begin{figure}[!h]
\centering 
% \rotatebox{90}{%

%  \includegraphics[width=1.3\textwidth]{Adobe_fig1_V3.png} 
%  }
 \caption{\doublespacing
Functionality of \pkg{metaDigitise}. Using the iris dataset in R, digitisation of different plot types, A) mean/error plot, B) box plot, C) histogram and D) scatter plot, is shown in \pkg{metaDigitise} (left) compared with other common softwares (right). A) and B) are plotted with the whole dataset, C) is just the data for the species \textit{setosa} and D) a subset from all three species. Notable functions of metaDigitise are listed in the center. Other software also perform points 3 and 4 (see Table \ref{tab:comparison}), although these functions are more developed in \pkg{metaDigitise}. As shown on the left hand side of the figure, \pkg{metaDigitise} clearly displays the stages of the digitisation to aid the transparency of the process, and returns concatenated summary data for all images.
%  All figures are clearly labelled at the top to remind users of the filename and plot type. This reduces errors throughout the digitisation process. Names of the variables and calibration (in blue) are plotted alongside the digitised points. In A) and B), user entered group names and sample sizes are displayed beside the relevant points. In C) the names and sample sizes for each group are shown below the figure.
}
\label{fig:all_extract}
\end{figure}

>>>>

In many different contexts, researchers make use of data presented in primary literature. In the fields of ecology and evolution (E\&E), this data is most commonly used for comparative and meta-analyses. Meta-analysis in particular is a rapidly expanding methodology in E\&E, not only in terms of the number of meta-analyses (in plant ecology alone the yearly number of published meta-analyses doubled from 2006 to 2012 (20-40); \cite{Koricheva2014}), but also in terms of their size (a recent meta-analysis, for example, included 6440 effect sizes from 175 publications; \cite{Noble2018}). The integrative nature of these studies makes them extremely important to the progression of E\&E \citep{Gurevitch2018}. These techniques rely heavily on descriptive statistics (e.g. means, standard deviations (SD), sample sizes, correlation coefficients) extracted from primary literature. As well as being presented in the text or tables of research papers, descriptive statistics are frequently presented in figures and so need to be manually extracted using digitising programs.

Although there are several tools that extract data from figures, including both standalone programs and \proglang{R} packages (reviewed in Table \ref{tab:comparison}), these tools do not cater to the general needs of meta-analysts for four main reasons (here we focus on meta-analysis, although many  points apply to extraction for comparative analysis). 
%
 First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}.
%
 Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later, typically using different software. % Given the ubiquity of the R platform in E\&E, and that the most popular meta-analysis software in E\&E (e.g., metafor and MCMCglmm) are available as R packages, it is highly likely to be used for some (if not all) stages of the research synthesis process. It would therefore be very beneficial to have comprehensive, and robust, digitisation capabilities in R that provides users with lots of flexibility, and makes the process of figure extraction more transparent and easier to reproduce.
%
 Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images and export the resulting digitised data into individual files one-by-one. These data often subsequently need to be imported and edited using different software. 
%
 Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently, a large amount of additional data manipulation is required, that is different across plots types. For example, in E\&E data are commonly presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented. Histograms and box plots are also frequently used in E\&E to presented data, and whilst their downsream calculations are even more laborious, there are few (if any; see Table \ref{tab:comparison}) tools to extract data from these plot types.

Data extraction from figures is therefore a time-consuming process as existing software does not provide an optimized, reproducible research pipeline to facilitate data extraction and editing. Given the ubiquity of the R platform in E\&E, and that it hosts the most popular meta-analysis software in E\&E (e.g., metafor \citep{Viechtbauer2010} and MCMCglmm \citep{Hadfield2010b}), it is highly likely to be used for some (if not all) stages of the research synthesis process. It is therefore important to have comprehensive, robust and flexible digitisation capabilities in R to make the process of figure extraction more streamline,  transparent and easier to reproduce. Here, we present an interactive \proglang{R} package, \pkg{metaDigitise} (available on CRAN), which is designed for large scale, reproducible data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools to extract data from common plot types in E\&E (mean/error plots, box plots, scatter plots and histograms, see Figure \ref{fig:all_extract}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. The necessary calculations are carried out on calibrated data immediately after extraction so that comparable descriptive statistics can be obtained quickly. Summary data from multiple figures is returned into a single data frame which can be can easily exported or used in downstream analysis within \proglang{R}. Completed digitisations are automatically saved for each figure, meaning users can redraw their digitisations (along with metadata) on figures, make corrections and access calibration and processed (i.e., summarised) data. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analyses to be updated more efficiently.

@@

\section{\pkg{metaDigitise} and Reproducibility}
The \pkg{metaDigitise} package has one main function, \fct{metaDigitise}, which interactively takes the user through the process of extracting data from figures (see Supplementary Material S1 for a full tutorial). Running \fct{metaDigitise} presents the user with three options; `Process new images', `Import existing data' or `Edit existing data', which can be used during and after digitisation to execute a range of functions (see Figure \ref{fig:all_extract} – `Processing images' is discussed in Section 3, and `Editing' and `Importing' in Section 4). \fct{metaDigitise} works on a directory containing images of figures copied from primary literature, in .png, .jpg, .tiff, .pdf format, specified to \fct{metaDigitise} through the \code{dir} argument. \fct{metaDigitise} recognizes all the images in the given directory and automatically imports them one-by-one, allowing the user to extract the relevant information about a figure as they go. Figures can be organised in different ways for a project, but we would recommend having all figures for one project in a single directory with an informative and unambiguous naming scheme (e.g. paper\_figure\_trait.png). This expedites digitisation by preventing users from having to constantly change directories and / or open new images.

@@

The complete digitisation process can be reproduced at a later stage, shared with collaborators and presented as supplementary materials for a publication, regardless of the computer it is run on. To update an analysis, new figures can simply be added to the directory and \fct{metaDigitise} run to incorporate the new data.

@@

\section{Image Processing}
 Selecting `Process New Images', after running \fct{metaDigitise}, starts the digitisation process on images within the directory that have not previously been digitised. For all plot types, \fct{metaDigitise} requires the user to calibrate the axes in the figure, by clicking on two known points on the axis in question, and entering the value of those points (Figure \ref{fig:all_extract}). \fct{metaDigitise} then calculates the value of any clicked points in terms of the figure axes. This is based on the calibration used in the \pkg{digitize} R package \citep{Poisot2011}. For mean/error and box plots, only the y-axis is calibrated (Figure \ref{fig:all_extract}), assuming the x-axis is redundant. For scatter plots and histograms both axes are calibrated (Figure \ref{fig:all_extract}).

Calibration of points in figures from older, scanned publications can be problematic, as the figures may not be perfectly orientated. \fct{metaDigitise} allows users to rotate the image (Figure S2A,B). Furthermore, mean/error plots, box plots and histograms, may be presented with horizontal bars. \fct{metaDigitise} assumes that bars are vertical, but allows the user to flip the image to make the bars are vertical (Figure S2C,D). \pkg{metaDigitise} also allows back calculation of data presented on log axes.

@@

\section{Importing and Editing Previously Digitised data}
\pkg{metaDigitise} is also able to re-import, edit and re-plot previously digitised figures. When running \fct{metaDigitise}, the user can choose to `Import existing data' , which returns previously digitised data, from a single figure or all figures. Alternately, the \fct{getExtracted} function returns the data from previous digitisations, but without user interaction, allowing easier integration into larger scripts. `Edit existing data' allows the user to re-plot or edit information for digitisations that have previously be done. Re-plotting digitisations with all metadata is an important reproducibility feature, as it allows users to see exactly what information has been extracted, as well as making it easy to spot and data extraction errors.

@@

\section{Software Validation}
In order to evaluate the consistency of digitisation with \pkg{metaDigitise} between users, fourteen people digitized sets of 14 identical images created from a simulated dataset (see supplements). We found no evidence for any inter-observer variability in digitisations for the mean (ICC = 0, 95\% CI = 0 to 0.029, \textit{p} $>$ 0.999), SD (ICC = 0, 95\% CI = 0 to 0.033, \textit{p} $>$ 0.999) or correlation coefficient (ICC = 0.053, 95\% CI = 0 to 0.296, \textit{p} = 0.377). There was little bias between digitised and true values, on average 1.63\% (mean = 0.02\%, SD = 4.9\%, \textit{r} = -0.03\%) and there were small absolute differences between digitised and true values, on average 2.18\% (mean = 0.40\%, SD = 5.81\%, \textit{r} = 0.33\%) across all three descriptive statistics. SD estimates from digitisations are clearly most error prone. The mean absolute differences for each plot type clearly show that this effect is driven by extraction from box plots and histograms (\% difference; box plot: 15.81, histogram: 5.21, mean/error: 1.50, scatter plot: 0.43). SD estimation from box plot descriptive statistics is known to be more error prone, especially at small sample sizes \citep{Wan2014}.

We also used simulated data to test the accuracy of digitisations with respect to known values (see supplements). \pkg{metaDigitise} was very accurate at matching clicked points to their true values essentially being perfectly correlated with the true simulated data for both the \textit{x}-variable (Pearson's correlation; $r > 0.999$, \textit{t} = 2137.4, \textit{df} = 78, $p < 0.001$) and \textit{y}-variable ($r > 0.999$, \textit{t} = 1897.8, \textit{df} = 78, $p < 0.001$) in scatterplots.

@@

In contrast to some other packages \pkg{metaDigitise} does not extract lines from figures. Although line extraction is not generally necessary in meta-analysis and comparative analysis, outside of these fields researchers may need to extract parameters of a line from a figure. Should a user like to extract lines with \pkg{metaDigitise}, we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points (accessed by choosing to return calibrated rather than summary data) to estimate the parameters needed.

@@

\begin{figure}[!h]
\centering 
\rotatebox{90}{%
  \includegraphics[width=1.3\textwidth]{Adobe_fig1_V3.png} 
 }
 \caption{%\doublespacing
Functionality of \pkg{metaDigitise}. Using the iris dataset in R, digitisation of different plot types, A) mean/error plot, B) box plot, C) histogram and D) scatter plot, is shown in \pkg{metaDigitise} (left) compared with other common softwares (right). A) and B) are plotted with the whole dataset, C) is just the data for the species \textit{setosa} and D) a subset from all three species. Notable functions of metaDigitise are listed in the center. Other software also perform points 3 and 4 (see Table \ref{tab:comparison}), although these functions are more developed in \pkg{metaDigitise}. As shown on the left hand side of the figure, \pkg{metaDigitise} clearly displays the stages of the digitisation to aid the transparency of the process, and returns concatenated summary data for all images.
}
\label{fig:all_extract}
\end{figure}
